---
title: "ProblemSet3"
author: "Ke-li Chiu & Diego Mamanche Castellanos"
date: "06/02/2020"
output: pdf_document
abstract: "Education is a contributor to many beneficial socio-economic outcomes. We intend to exam the the relationship between academic majors and household income in Toronto's neighbourhoods—can we find higher or lower percentage population in certain academic majors in higher or lower income neighbourhoods? The result shows that higher income neighbourhoods also have higher percentage of residents with majors in business, The data set is retrieved from the package *City of Toronto Neighbourhood Profiles* from the year 2016. ###### We intend to use the comparison to bring awareness of inequality in income and education existing in the City of Toronto."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, include=FALSE, message=FALSE}
#Set up the environment
library(opendatatoronto)
library(dplyr)
library(tidyr)
library(tidyverse)
library(ggplot2)
library(knitr)
library(data.table) #Useful to transponse a dataframe
library(sf)
library(tmap)
library(tmaptools)
library(leaflet)
library(cluster)
```

# Introduction
Education is a contributor to many beneficial socio-economic outcomes. In this study, we intend to examine the relationship between household income and educational profile in terms of residents' academic majors in Toronto's 140 neighbourhoods. The majors are categorized as follows—
"education", "visual and performing arts and communications technologies," "humanities," "social and behavioural sciences and law," "business_management_and_public_administration," "physical and life sciences and technologies," "mathematics computer and information sciences," "architecture engineering and related technologies," "agriculture natural resources and conservation," "personal protective and transportation services" and "no postsecondary certificate diploma or degree." The question we want to answer is if we can find a higher population percentage in specific academic majors in higher-income neighbourhoods or vice versa? The question is relevant because the academic major is a significant factor for people's occupations and their associated earning; thus, it is deemed as an essential tool for social mobility.

```{r echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
#Set working Directory
#setwd("~/Experimental Design for Data Science/ProblemSet2")
# Search packages (this returns a table)
neighbourhood_packages <- search_packages("Neighbourhood")
# Filter the neighbourhood demographic dataset
neighbourhood_demographics_package<- neighbourhood_packages %>%
  filter(title == "Neighbourhood Profiles")
#Create main dataset for year 2016 
main_df_2016 <-  neighbourhood_demographics_package %>% # Start with the package 
  list_package_resources() %>% # List the resources in the package
  filter(name == "neighbourhood-profiles-2016-csv") %>% # Only keep the resource we want 
  get_resource()
### GEO-LOCATION DATASET TORONTO BY NEIGHBOURHOODS ###
# get package
package <- show_package("4def3f65-2a65-4a4f-83c4-b2a4aed72d46")
package
# get all resources for this package
resources <- list_package_resources("4def3f65-2a65-4a4f-83c4-b2a4aed72d46")
# identify datastore resources; by default, Toronto Open Data sets datastore resource 
#format to CSV for non-geospatial and GeoJSON for geospatial resources
datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
# load the first datastore resource as a sample
geo_data <- filter(datastore_resources, row_number()==1) %>% get_resource()
```


# Dataset and Method Description
City of Toronto Neighbourhood Profiles data set is sourced from several Census tables released by Statistics Canada every five years. The dataset uses this Census data to provide a portrait of the demographic, social and economic characteristics of the people and households in each City of Toronto neighbourhood. Each data point in this file is presented for the City's 140 neighbourhoods, as well as for the City of Toronto as a whole. The data set consists of 144 columns for neighbourhoods and 1537 rows for characteristics of the neighbourhoods. Data cleaning and reshaping are performed to have a new data frame that has 14 characteristics regarding topics of income as columns and education and the 140 neighbourhoods as rows.ing among academic fields.

```{r echo=FALSE, include=TRUE, message=FALSE, warning=FALSE}
##### Distribution by age 2016 ####

#Select sheet 2016
clean_header_df_2016 <- main_df_2016
#Clean header using janitor
clean_header_df_2016 <- janitor::clean_names(clean_header_df_2016)
#Select only Population by Age characteristics from the main dataset 
age_df_2016 <- filter(clean_header_df_2016, category == "Population" & 
                              topic == "Age characteristics")


#Remove rows 1 to 6. It contains general information not related with the subject under analysis
age_df_2016 <- age_df_2016[7:nrow(age_df_2016),]

#Remove first 4 columns and column "city_of_toronto" 
age_df_2016 <- age_df_2016[,-c(1:4,6)]


#Select only the characters after the word "to" or those finishing in "over"
age_df_2016 <- extract(age_df_2016, characteristic, into = "age_range" , regex = "([to][ 0-9]+[ years]+$|[0-9 ]+[years and over]+$)", remove = FALSE)

#Extract only the numbers 
age_df_2016 <- extract(age_df_2016, age_range, into = "age_range" , regex = "([0-9]+)", remove = TRUE)

#Extract biosex in a separate column
age_df_2016 <- extract(age_df_2016, characteristic, into = "biosex" , regex = "(^[A-z]+[^:])", remove = TRUE) #The remove = False keeps the older column



#Change class of columns containing numbers as.numeric 
age_df_2016[,3:length(age_df_2016)] <- 
  sapply(age_df_2016[,3:length(age_df_2016)], function(x) as.numeric(x))

#Create new dataframe
scaled_age_df_2016 <- age_df_2016
#Scale dataframe by column
scaled_age_df_2016 <-
  scaled_age_df_2016 %>%
    janitor::adorn_percentages(denominator = "col")
#Multiply by 100
scaled_age_df_2016[,3:length(scaled_age_df_2016)] <- 
  sapply(scaled_age_df_2016[,3:length(scaled_age_df_2016)], function(x) (x*100))
#Round new scales to 2 digits
scaled_age_df_2016 <-
  scaled_age_df_2016 %>%
    janitor::adorn_rounding(digits = 2)
#Transform the result to dataframe
scaled_age_df_2016 <- as.data.frame(scaled_age_df_2016)


#age_df_2016_reshaped <- transpose(scaled_age_df_2016, make.names=1, keep.names="neighbourhoods")

##Reshape all neighbourhood columns into one column called variable and the values
scaled_age_df_2016 <- melt(scaled_age_df_2016, measure = 3:length(scaled_age_df_2016))


### Male by age
age_male_2016 <- filter(scaled_age_df_2016, biosex == "Male")

age_male_2016 <- reshape(age_male_2016[,-1], idvar="variable", timevar="age_range", direction="wide")

#colnames(scaled_age_df_2016) <- c("neighbourhoods", "age_range", "male", "female") 


### Female by age
age_female_2016 <- filter(scaled_age_df_2016, biosex == "Female")

age_female_2016 <- reshape(age_female_2016[,-1], idvar="variable", timevar="age_range", direction="wide")

scaled_age_df_2016 <- merge(age_male_2016, age_female_2016, by="variable")

colnames(scaled_age_df_2016) <- gsub("variable", "neighbourhoods", colnames(scaled_age_df_2016))
colnames(scaled_age_df_2016) <- gsub("\\.", "_", colnames(scaled_age_df_2016))

colnames(scaled_age_df_2016) <- gsub("value_", "", colnames(scaled_age_df_2016))

colnames(scaled_age_df_2016) <- gsub("x", "male", colnames(scaled_age_df_2016))
colnames(scaled_age_df_2016) <- gsub("y", "female", colnames(scaled_age_df_2016))


head(scaled_age_df_2016)
```


```{r echo=FALSE, include=TRUE, message=FALSE}
##### Distribution by income 2016####

#Select sheet 2016
clean_income_df_2016 <- main_df_2016
#Clean header using janitor
clean_income_df_2016 <- janitor::clean_names(clean_income_df_2016)


#Select only Population by Income characteristics from the main dataset 
income_df_2016 <- filter(clean_income_df_2016, category == "Income" & 
                              topic == "Income of households in 2015")

#Remove columns 1 to 4 and column "city_of_toronto" 
income_df_2016 <- income_df_2016[,-c(1:4,6)]

#Remove rows not related to total income by intervals
income_df_2016 <- income_df_2016[-c(2:6,8:13,15:20,22:23,38,40:59),]

#Extract Income in a separate column
income_df_2016 <- extract(income_df_2016, characteristic, into = "income_range" , 
                       regex = "([0-9,]+$|[0-9, and over]+$)", remove = TRUE) #The remove = False keeps the older column


#Extract Income in a separate column
income_df_2016 <- extract(income_df_2016, income_range, into = "income_range" , 
                       regex = "([,0-9]+)", remove = TRUE) #The remove = False keeps the older column



#Remove "," in all numerical variables
for (i in colnames(income_df_2016[,2:length(income_df_2016)])) {
  
  eval(parse(text= paste("income_df_2016$", i, "<- ",   'gsub(",", "", income_df_2016$', i, ")", sep = "")))
  
}

#Change class of columns containing numbers as.numeric 
income_df_2016[,2:length(income_df_2016)] <- 
  sapply(income_df_2016[,2:length(income_df_2016)], function(x) as.numeric(x))

#Create new dataframe
scaled_income_df_2016 <- income_df_2016
#Scale dataframe by column
scaled_income_df_2016 <-
  scaled_income_df_2016 %>%
    janitor::adorn_percentages(denominator = "col")
#Multiply by 100
scaled_income_df_2016[,2:length(scaled_income_df_2016)] <- 
  sapply(scaled_income_df_2016[,2:length(scaled_income_df_2016)], function(x) (x*100))
#Round new scales to 2 digits
scaled_income_df_2016 <-
  scaled_income_df_2016 %>%
    janitor::adorn_rounding(digits = 2)
#Transform the result to dataframe
scaled_income_df_2016 <- as.data.frame(scaled_income_df_2016)

inco_df_2016_reshaped <- transpose(scaled_income_df_2016, make.names=1, keep.names="neighbourhoods")



##Reshape all neighbourhood columns into one column called variable and the values
#scaled_income_df_2016 <- reshape2::melt(scaled_income_df_2016, measure = #2:length(scaled_income_df_2016))


head(inco_df_2016_reshaped)
```

```{r}
#### Hierarchical Clustering by Age ###

#Normalization
age_all_num <- scaled_age_df_2016[,-1]
mean_age_df_2016 <- apply(age_all_num,2,mean) 
sd_age_df_2016 <- apply(age_all_num,2,sd) 
norm_age_df_2016 <- scale(age_all_num,mean_age_df_2016, sd_age_df_2016)

#Calculate euclidean distance
age_distance <- dist(norm_age_df_2016, method = "euclidean")

#Cluster dendogram with Complete linkage
age_hclust <- hclust(age_distance)
plot(age_hclust, hang = -1)

#Cluster membership
age_membership <- cutree(age_hclust, k = 5)

#Cluster means
aggregate(age_all_num, list(age_membership), mean)
aggregate(scaled_age_df_2016[,-1], list(age_membership), mean)

#Silhouette Plot
plot(silhouette(cutree(age_hclust, k = 5), age_distance))

#Scree plot
age_ww_sum_sqrs <- (nrow(age_all_num) -1)*sum(apply(age_all_num,2,mean))
for (i in 2:20) age_ww_sum_sqrs[i] <- sum(kmeans(age_all_num, centers = i)$withinss)
plot(1:20, age_ww_sum_sqrs, type="b", xlab="number of clusters", ylab = "Within group SS")

```

```{r}
#### Hierarchical Clustering by Income ###

#Normalization
income_all_num <- inco_df_2016_reshaped[,-1]
mean_income_df_2016 <- apply(income_all_num,2,mean) 
sd_income_df_2016 <- apply(income_all_num,2,sd) 
norm_income_df_2016 <- scale(income_all_num,mean_income_df_2016, sd_income_df_2016)

#Calculate euclidean distance
income_distance <- dist(norm_income_df_2016, method = "euclidean")

#Cluster dendogram with Complete linkincome
income_hclust <- hclust(income_distance)
plot(income_hclust, hang = -1)

#Cluster membership
income_membership <- cutree(income_hclust, k = 5)

#Cluster means
aggregate(income_all_num, list(income_membership), mean)
aggregate(inco_df_2016_reshaped[,-1], list(income_membership), mean)

#Silhouette Plot
plot(silhouette(cutree(income_hclust, k = 5), income_distance))

#Scree plot
income_ww_sum_sqrs <- (nrow(income_all_num) -1)*sum(apply(income_all_num,2,mean))
for (i in 2:20) income_ww_sum_sqrs[i] <- sum(kmeans(income_all_num, centers = i)$withinss)
plot(1:20, income_ww_sum_sqrs, type="b", xlab="number of clusters", ylab = "Within group SS")


```

```{r}
### K-means clustering by age ###
age_kmeans_clust <- kmeans(norm_age_df_2016,4)
age_kmeans_clust

```

```{r}
### K-means clustering by age ###
income_kmeans_clust <- kmeans(norm_income_df_2016,3)
income_kmeans_clust

```

```{r}
### Cleaning geo-location dataset
clean_geo_data <- janitor::clean_names(geo_data)
clean_geo_data <- extract(clean_geo_data, area_name, into = "neighbourhoods" , regex = "([^(0-9)]+)")
clean_geo_data["neighbourhoods"] <- 
  janitor::make_clean_names(as.matrix(clean_geo_data["neighbourhoods"]))
clean_geo_data <- janitor::clean_names(clean_geo_data)
clean_geo_data <- select(clean_geo_data, neighbourhoods, longitude, latitude, geometry)
filter(clean_geo_data, neighbourhoods %in% c("mimico","weston_pellam_park"))
clean_geo_data["neighbourhoods"][c(17,67),] <- c("mimico_includes_humber_bay_shores","weston_pelham_park")
clean_geo_data
```
```

```{r}
### Merge age hierarchical cluster and Kmeans classification with the dataset ###

#Merge Kmeans
df_kmeans_age <- mutate(scaled_age_df_2016, kmeans = age_kmeans_clust$cluster)

merge_age_clust <- merge(df_kmeans_age, clean_geo_data, by = 'neighbourhoods')

#Merge Hclustering
hcluster <- cutree(plot_data_comp, k=4)
hcluster <- as.data.frame(hcluster)
hcluster$hcluster
df_w_vector <- mutate(df_cleaned, hcluster = hcluster$hcluster)



```











\pagebreak
# Appendix A
```{r ref.label=knitr::all_labels(), echo = T, eval = F, tidy=TRUE, tidy.opts=list(width.cutoff=60), results='asis', width.cutoff=60}
```

\pagebreak
# References


Open Data Dataset. (n.d.). Retrieved January 30, 2020, from https://open.toronto.ca/dataset/
wellbeing-toronto-demographics-nhs-indicators/

R Core Team (2019). R: A language and environment for statistical computing. R
Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.

H. Wickham. ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York, 2016.

Sharla Gelfand (2019). opendatatoronto: Access the City of Toronto Open Data Portal. R
package version 0.1.1. https://CRAN.R-project.org/package=opendatatoronto
  
Hadley Wickham, Romain François, Lionel Henry and Kirill Müller (2019). dplyr: A Grammar
of Data Manipulation. R package version 0.8.3. https://CRAN.R-project.org/package=dplyr

Hadley Wickham and Lionel Henry (2019). tidyr: Tidy Messy Data. R package version 1.0.0.
https://CRAN.R-project.org/package=tidyr

Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software,
4(43), 1686, https://doi.org/10.21105/joss.01686

Yihui Xie (2019). knitr: A General-Purpose Package for Dynamic Report Generation in R. R
package version 1.25.

Yihui Xie (2015) Dynamic Documents with R and knitr. 2nd edition. Chapman and Hall/CRC.
ISBN 978-1498716963

Yihui Xie (2014) knitr: A Comprehensive Tool for Reproducible Research in R. In Victoria
Stodden, Friedrich Leisch and Roger D. Peng, editors, Implementing Reproducible

Matt Dowle and Arun Srinivasan (2019). data.table: Extension of `data.frame`. R package
version 1.12.6. https://CRAN.R-project.org/package=data.table
Computational Research. Chapman and Hall/CRC. ISBN 978-1466561595

Pebesma, E., 2018. Simple Features for R: Standardized Support for Spatial Vector Data.
The R Journal 10 (1), 439-446, https://doi.org/10.32614/RJ-2018-009

Tennekes M (2018). “tmap: Thematic Maps in R.” _Journal of Statistical Software_, *84*(6),
1-39. doi: 10.18637/jss.v084.i06 (URL: https://doi.org/10.18637/jss.v084.i06).

Martijn Tennekes (2019). tmaptools: Thematic Map Tools. R package version 2.0-2.
https://CRAN.R-project.org/package=tmaptools

Joe Cheng, Bhaskar Karambelkar and Yihui Xie (2019). leaflet: Create Interactive Web
Maps with the JavaScript 'Leaflet' Library. R package version 2.0.3.
https://CRAN.R-project.org/package=leaflet

Maechler, M., Rousseeuw, P., Struyf, A., Hubert, M., Hornik,
K.(2019).  cluster: Cluster Analysis Basics and Extensions. R package
version 2.1.0.